---
title: "In class Exercise 6"
description: |
  In this in class ex 6, we will explore geo-spatial analytics.
  
author:
  - name: Hulwana Saifulzaman 
    url: https://www.linkedin.com/in/hulwana-saifulzaman/
    affiliation: SMU, Master of IT in Business
    affiliation_url: https://scis.smu.edu.sg/master-it-business
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

## 1. Load the packages
```{r}
packages = c('tidyverse','sf','tmap', 'sftime', 'lubridate', 'clock', 'rmarkdown')

for (p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p,character.only = T)
}
```

## 2. Importing wkt data

```{r}
schools<- read_sf("data/wkt/Schools.csv", options = "GEOM_POSSIBLE_NAMES=location")

pubs <- read_sf("data/wkt/Pubs.csv", options = "GEOM_POSSIBLE_NAMES=location")

apartments <- read_sf("data/wkt/Apartments.csv", options = "GEOM_POSSIBLE_NAMES=location")

buildings <- read_sf("data/wkt/Buildings.csv", options = "GEOM_POSSIBLE_NAMES=location")

employers <- read_sf("data/wkt/Employers.csv", options = "GEOM_POSSIBLE_NAMES=location")

restaurants <- read_sf("data/wkt/Restaurants.csv", options = "GEOM_POSSIBLE_NAMES=location")

bldgs <- read_sf("data/wkt/Buildings.csv", options = "GEOM_POSSIBLE_NAMES=location")
```

## 3. Review loaded data

### 3.1. Schools

```{r}
print(schools)
```

### 3.2. Buildings

```{r}
print(buildings )
```

## 4. Plot point symbol map

#### tmap_mode("view") to make interactive
#### tmap_mode("plot")  to make it static, can put at the end of the code to switch off the interactive
##### try not too have too many interactive map, as it might take a lot of time to render

#### tm_shape: to define the data
#### based on the geometry type of the data use tm_polygons (for polygons) / tm_dots (for points)
#### tm_polygons: col - the colour of the polygon
##### when you click on it, it wil show the data

##### sequence is important

```{r}
tmap_mode("view")
tm_shape(buildings)+
tm_polygons(col = "grey60",
           size = 1,
           border.col = "black",
           border.lwd = 1)
tmap_mode("plot")
```

##### sequence is important, plot area > line > dot, so it wont be covered by the bigger plot
##### if you layer the employer first then the building, then you wont see the employer

```{r eval=FALSE}
tmap_mode("view")
tm_shape(buildings)+
tm_polygons(col = "grey60",
           size = 1,
           border.col = "black",
           border.lwd = 1) +
tm_shape(employers) +
  tm_dots(col = "red")
tmap_mode("plot")
```


```{r}
tmap_mode("view")
tm_shape(buildings)+
tm_polygons(col = "grey60",
           size = 1,
           border.col = "black",
           border.lwd = 1) +
tm_shape(employers) +
  tm_dots(col = "red") +
tm_shape(apartments) +
  tm_dots(col = "lightblue") +
tm_shape(pubs) +
  tm_dots(col = "green") +
tm_shape(restaurants) +
  tm_dots(col = "blue") +
tm_shape(schools) +
  tm_dots(col = "yellow")
tmap_mode("plot")
```


## ParticipantLog

```{r echo =FALSE, eval=FALSE}
logs <- read_sf("data/wkt/ParticipantStatusLogs1.csv", options = "GEOM_POSSIBLE_NAMES=currentLocation")
```


## Data prep

```{r eval=FALSE}
logs_selected <- logs %>% mutate(Timestamp = date_time_parse(timestamp,
                                                             zone = "",
                                                             format = "%Y-%m-%dT%H:%M:%S")) %>% 
  mutate(day = get_day(Timestamp)) %>% 
  filter(currentMode == "Transport")
```

## to save it in rds form

```{r echo =FALSE, eval=FALSE}
write_rds(logs_selected, 
          "data/rds/logs_selected.rds")
```

##  load the rds
```{r}
logs_selected <- read_rds("data/rds/logs_selected.rds")
```

## Computing hexagons

```{r}

hex <- st_make_grid(buildings,
                    cellsize = 100,
                    square = FALSE) %>%
  st_sf() %>%
  rowid_to_column('hex_id')

plot(hex)
```


```{r}
points_in_hex <- st_join(logs_selected, 
                        hex, 
                        join=st_within) %>%
  st_set_geometry(NULL) %>%
  count(name='pointCount', hex_id)
head(points_in_hex)
```

## relational join 

```{r}
hex_combined <- hex %>%
  left_join(points_in_hex, 
            by = 'hex_id') %>%
  replace(is.na(.), 0)
```

## plot hex map

```{r}
tm_shape(hex_combined %>%
           filter(pointCount > 0))+
  tm_fill("pointCount",
          n = 8,
          style = "quantile") +
  tm_borders(alpha = 0.1)
```

## movement path

```{r}
logs_path <- logs_selected %>%
  group_by(participantId, day) %>%
  summarize(m = mean(Timestamp), 
            do_union=FALSE) %>%
  st_cast("LINESTRING")
```

